Blockchain System

### 2020-12-21 papers/execution-model/model.tex
intermediate; ongoing discussion

\begin{definition}%[Account]
  An implicit \emph{account} is a tuple $\langle
  %\ALS,
  \PAK, \PUK
  %, \PKH
  \rangle $, where
  $\PAK$ is its private key and $\PUK$ is its public key.
\end{definition}

@@

\begin{definition}%[Program]
\emph{Expressions} are defined by the following grammar:
\begin{align*}
  \CONSTANT & ::= \ADDR \mid \OPH \mid \PUH \mid\PUK \mid \CODE \mid
              \NTEZ \mid \SUNIT \mid \FALSE \mid \TRUE \\
  \STATUS & ::= \STATUSPENDING \mid \STATUSINCLUDING (\dots) \mid
            \STATUSTIMEOUT \\
  \ERROR &::= \ERRPRG \mid \ERRBAL \mid \ERRCOUNT \mid \ERRFEE \mid
           \ERRPUK \mid \ERRPUH \mid \ERRARG\\
	\EXPR & :: =  \CONSTANT \mid \STATUS \mid \ERROR \mid \VARIABLE \mid \lambda \VARIABLE. \EXPR\ 
	\mid \EXPR\EXPR \mid (\EXPR,\EXPR) \mid \FST\ \EXPR \mid \SND\ \EXPR 	%\mid \NI \EI
	\\&\mid \EXPR\ \PLUS\ \EXPR\ \mid \EXPR\ \MINUS\ \EXPR\ \mid \EXPR\ \EQUAL\ \EXPR\ \mid \EXPR\ \LESS\ \EXPR
	\\&\mid \EXPR\ \AND\ \EXPR\ \mid \EXPR\ \OR\ \EXPR\ \mid \NOT\ \EXPR
  \\&\mid \NIL\ \mid \CONS\ \EXPR\ \EXPR \mid \LEFT\ \EXPR \mid\RIGHT\
  \EXPR \mid \SOME\ \EXPR \mid \NONE
  \\& \mid \RAISE\ \EXPR \mid \TRY\ \EXPR\ \EXCEPT\ \EXPR
  \\& \mid \MATCH\ \EXPR\ \WITH\ \PATTERN\to\EXPR \dots
  \\& \mid \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR
  \\& \mid \ORIGINATE\EXPR\EXPR\EXPR\EXPR\EXPR
  \\&\mid \QOP\  \EXPR\\
  \QOP & ::= balance \mid status \mid storage \mid code \mid counter
\end{align*}
\end{definition}

Example code
\begin{align*}
  \OPH &=  \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR \\
       & \MATCH\ {status\ \OPH}\ \WITH \\
       & \mid \STATUSPENDING \to \textrm{not yet included, continue waiting} \\
       & \mid \STATUSINCLUDING (n) \to \textrm{included with $n$
         blocks confirmation}\\
       & \mid \STATUSTIMEOUT \to \textrm{lost, try again}
\end{align*}
Grammar of types
\begin{align*}
  \TYPE & ::= \ADDR \mid \OPH \mid \PUH \mid \PUK \mid \CODE\ \TYPE \mid
          \TINT \mid \TUNIT \mid \TBOOL \mid \TSTRING \mid \TSTATUS \\
  & \mid \TYPE\to\TYPE \mid \TPAIR\ \TYPE\ \TYPE \mid \TLIST\ \TYPE
    \mid \TSUM\ \TYPE\ \TYPE \mid \TOPTION\ \TYPE \mid \TEXCEPTION
\end{align*}

\begin{definition}%[State of a node]
  The \emph{state of a node} is a tuple
  $\NODE = [ \EXPRS, \ACCOUNTS, \OPERATIONS]$, where $\EXPRS$ is the set of the programs, $\ACCOUNTS$ is a set of
  accounts, and
  $\OPERATIONS$ is a set of operations.
\end{definition}

@@

\begin{definition}%[Manager]
A \emph{manager} manages a single implicit account. It is represented by a
tuple $\langle  \PUK, \BAL, \COU \rangle $, where  $\PUK$ is the
public key of an account, $\BAL$ is its
balance and $\COU$ is its counter whose form is a value-flag pair $(n, b)$, where $n$, which is called the value of the counter,  is a natural number and ${b}$, which is called the flag of the counter, is a boolean value.  
\end{definition}

\begin{definition}%[Contractor]
  A \emph{contractor} manages a smart contract. It is
  represented by a tuple
  $\langle \PUH, \BAL, \CODE, \STORAGE \rangle $, where $\PUH$ is the
  public hash of the contract, $\BAL$ is its current balance, $\CODE$ is its
  code, and $\STORAGE$ is its current storage.
\end{definition}

When an operation is injected in a node, it enters a \emph{pending pool}
(and is called a \emph{pending operation}).

 \begin{definition}%[Pending operation]
A \emph{pending operation} is a tuple  $\langle  \OP, \OPH, \TIME
\rangle $, where $\OP$ is an operation, $\OPH$ is the operation hash,
and $\TIME$ is the time when the operation was injected. 
\end{definition}

After some time, a pending operation may be included in the blockchain as an accepted operation.

 \begin{definition}%[Accepted operation]
An \emph{accepted operation}  is a tuple $\langle  \OP, \OPH, \TIME, \TIME' \rangle
$, where $\OP$ is an operation, $\OPH$ is the operation hash, $\TIME$ is the time when the operation was injected, and
$\TIME'$ is the time when it was included in the blockchain. 
\end{definition}

\begin{definition}%[Blockchain]
  The \emph{state of a blockchain} is a tuple
  $[\PENDING, \ACCEPTED, \MANAGERS, \CONTRACTORS, \TIME]$ where
  $\PENDING$ is a set of pending operations,  $\ACCEPTED$ is a set
  of accepted operations, $\MANAGERS$ is a set of managers,
  $\CONTRACTORS$ is a set of contractors, and $\TIME$ is the current time. 
\end{definition}

A \emph{pending operation} is a tuple
$\langle  \OP, \OPH, \TIME, \STATUSPENDING\rangle $,
where $\OP$ is an operation, $\OPH$ is the operation hash,
and $\TIME$ is the time when the operation was injected.
Once it is accepted, it changes to an \emph{accepted operation}:
$\langle  \OP, \OPH, \TIME, \STATUSINCLUDED\ \TIME'\rangle $, where
$\TIME'$ is the  when it was included in the blockchain.

\begin{definition}%[Blockchain system]
A \emph{blockchain configuration} is a pair
$ \NODE \| \BLOCKCHAIN$ where
\fxnote[inline]{Should the model contain multiple nodes?}
\begin{enumerate}
\item $\NODE = [\ACCOUNTS, \OPERATIONS]$ is the state of a node, and
\item $\BLOCKCHAIN = [\PENDING, \ACCEPTED, \MANAGERS, \CONTRACTORS,
  \TIME]$ is the state of a blockchain  
  such that
  \begin{itemize}
  \item
    $\forall c \in \ACCOUNTS \implies \exists k\in \MANAGERS, k.\PUK =
    c.\PUK$;
  \item 
  \end{itemize}
\end{enumerate}
\end{definition}

>>>>

\begin{definition}%[Account]
  An implicit \emph{account} is represented by a key pair $\langle
  %\ALS,
  \PAK, \PUK
  %, \PKH
  \rangle $, where
  $\PAK$ is its private key and $\PUK$ is its public key.
\end{definition}

@@

\begin{definition}%[Program]
\emph{Expressions} are defined by the following grammar:
\begin{align*}
  \CONSTANT & ::= \ADDR \mid \OPH \mid \PUH \mid\PUK \mid \CODE \mid
              \NTEZ \mid \SUNIT \mid \FALSE \mid \TRUE \\
  \STATUS & ::= \STATUSPENDING \mid \STATUSINCLUDING (\dots) \mid
            \STATUSTIMEOUT \\
  \ERROR &::= \ERRPRG \mid \ERRBAL \mid \ERRCOUNT \mid \ERRFEE \mid
           \ERRPUK \mid \ERRPUH \mid \ERRARG\\
	\EXPR & :: =  \CONSTANT \mid \STATUS \mid \ERROR \mid \VARIABLE \mid \lambda \VARIABLE. \EXPR\ 
	\mid \EXPR\EXPR \mid (\EXPR,\EXPR) \mid \FST\ \EXPR \mid \SND\ \EXPR 	%\mid \NI \EI
	\\&\mid \EXPR\ \PLUS\ \EXPR\ \mid \EXPR\ \MINUS\ \EXPR\ \mid \EXPR\ \EQUAL\ \EXPR\ \mid \EXPR\ \LESS\ \EXPR
	\\&\mid \EXPR\ \AND\ \EXPR\ \mid \EXPR\ \OR\ \EXPR\ \mid \NOT\ \EXPR
  \\&\mid \NIL\ \mid \CONS\ \EXPR\ \EXPR \mid \LEFT\ \EXPR \mid\RIGHT\
  \EXPR \mid \SOME\ \EXPR \mid \NONE
  \\& \mid \RAISE\ \EXPR \mid \TRY\ \EXPR\ \EXCEPT\ \EXPR
  \\& \mid \MATCH\ \EXPR\ \WITH\ \PATTERN\to\EXPR \dots
  \\& \mid \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR
  \\& \mid \ORIGINATE\EXPR\EXPR\EXPR\EXPR\EXPR
  \\&\mid \QOP\  \EXPR\\
  \QOP & ::= balance \mid status \mid storage \mid code \mid counter
\end{align*}
We let $\EXPRS$ range over sets of expressions.

We let $\ECS$ range over evaluation contexts. They are defined as
follows.
\begin{align*}
  \EC{} & :: = \SC[\overline\VAL\ \EC{}\ \overline\EXPR] \mid \RAISE\ \EC{} \mid \TRY\ \EC{}\ \EXCEPT\ \EXPR
  \mid \MATCH\ \EC{}\ \WITH\ \PATTERN\to\EXPR \dots
\end{align*}
Here $\SC$ ranges over the remaining syntactic constructors, which are
treated uniformly: evaluation proceeds from left to right.
\begin{align*}
  \VAL & ::= \CONSTANT \mid \STATUS \mid \ERROR \mid \lambda x.\EXPR \mid(\VAL, \VAL) \mid
         \NIL \mid \CONS\ \VAL\ \VAL \mid \LEFT\ \VAL \mid \RIGHT\
         \VAL \mid \SOME\ \VAL\mid \NONE
\end{align*}
\end{definition}

Example pseudo code
\begin{align*}
  \OPH &=  \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR \\
       & \MATCH\ {status\ \OPH}\ \WITH \\
       & \mid \STATUSPENDING \to \textrm{not yet included, continue waiting} \\
       & \mid \STATUSINCLUDING (n) \to \textrm{included with $n$
         blocks confirmation}\\
       & \mid \STATUSTIMEOUT \to \textrm{lost, try again}
\end{align*}
Grammar of types
\begin{align*}
  \TYPE & ::= \TADDR \mid \TOPH \mid \TPUH \mid \TPUK \mid \TCODE\ \TYPE \mid
          \TINT \mid \TUNIT \mid \TBOOL \mid \TSTRING \mid \TSTATUS \\
  & \mid \TYPE\to\TYPE \mid \TPAIR\ \TYPE\ \TYPE \mid \TLIST\ \TYPE
    \mid \TSUM\ \TYPE\ \TYPE \mid \TOPTION\ \TYPE \mid \TEXCEPTION
\end{align*}

\begin{definition}%[State of a node]
  The \emph{state of a node} is a tuple
  $\NODE = [ \EXPRS, \ACCOUNTS, \OPERATIONS]$, where $\EXPRS$ is a
  set of programs, $\ACCOUNTS \subseteq \TPAK \times \TPUK$  is a set of
  implicit accounts, and
  $\OPERATIONS \subseteq \ECS \times \TOP $ is a set of pairs of
  expression contexts and operations.
\end{definition}

@@

\begin{definition}%[Manager]
A \emph{manager} manages a single implicit account. Managers are
represented by a partial mapping $\MANAGERS : \TPUK \partialto \TBAL
\times \TCOU$. If $\MANAGERS (\PUK) = \Angle{\BAL, \COU}$ is defined, then  $\PUK$ is the
public key of an account, $\BAL$ is its
balance and $\COU$ is its counter whose form is a value-flag pair
$(n,b) \in \Nat\times\TBOOL$, where $n$ is the value of the counter
and ${b}$ is its flag.   
\end{definition}

\begin{definition}%[Contractor]
  A \emph{contractor} manages a smart contract. Contractors are
  represented by a partial mapping $\CONTRACTORS : \TPUH \partialto
  \TBAL \times \TCODE \times \TSTORAGE$. If $\CONTRACTORS (\PUH) = \langle
  \BAL, \CODE, \STORAGE \rangle $ is defined, then $\PUH$ is the
  public hash of a contract, $\BAL$ is its current balance, $\CODE$ is its
  code, and $\STORAGE$ is its current storage.
\end{definition}

When an operation is injected in a node, it enters a \emph{pending
  pool}. Later on, such a pending operation is either dismissed or
promoted to an accepted operation.

 \begin{definition}%[Pending operation]
An \emph{operation} is represented by a tuple  $\langle  \OP, \TIME, \STATUS
\rangle $, where $\OP$ is an operation, $\TIME$ is the time when the
operation was injected, and $\STATUS$ is either $\STATUSPENDING$,
$\STATUSINCLUDING\ \TIME'$, or $\STATUSTIMEOUT$.  
\end{definition}
\begin{definition}
  The pool of operations is defined as a partial mapping from
  operation hashes $\OPH$:
  \begin{align*}
    \PENDING &= \TOPH \partialto \TOP\times \TTIME \times \TSTATUS \\
    \TSTATUS&= (\STATUSPENDING + \STATUSINCLUDED\ \TTIME + \STATUSTIMEOUT)
  \end{align*}
\end{definition}

A \emph{pending operation} is represented by
$\langle  \OP, \TIME, \STATUSPENDING\rangle $,
where $\OP$ is an operation, $\OPH$ is the operation hash,
and $\TIME$ is the time when the operation was injected.
Once it is accepted, it changes to an \emph{accepted operation}:
$\langle  \OP, \TIME, \STATUSINCLUDED\ \TIME'\rangle $, where
$\TIME'> \TIME$ is the  when the operation was included in the
blockchain.  The operation can also be dropped at any time, which is
represented by
$\langle  \OP, \TIME, \STATUSTIMEOUT\rangle $. There are several
causes for dropping, primarily timeout or overflow of the pending
pool which is limited in size in the implementation.

% After some time, a pending operation may be included in the blockchain as an accepted operation.

 % \begin{definition}%[Accepted operation]
% An \emph{accepted operation}  is a tuple $\langle  \OP, \OPH, \TIME, \TIME' \rangle
% $, where $\OP$ is an operation, $\OPH$ is the operation hash, $\TIME$ is the time when the operation was injected, and
% $\TIME'$ is the time when it was included in the blockchain. 
% \end{definition}

\begin{definition}%[Blockchain]
  The \emph{state of a blockchain} is a tuple
  $\BLOCKCHAIN = [\PENDING, \MANAGERS, \CONTRACTORS, \TIME]$ where
  $\PENDING$ is a pool of operations, $\MANAGERS$ is a map of managers,
  $\CONTRACTORS$ is a map of contractors, and $\TIME$ is the current time. 
\end{definition}

Properties of blockchain state transitions

If $[\PENDING, \MANAGERS, \CONTRACTORS, \TIME] \BlockTrans{}
[\PENDING', \MANAGERS', \CONTRACTORS', \TIME']$, then
\begin{itemize}
\item $\TIME \le \TIME'$
\item $\DOM (\PENDING) \subseteq \DOM(\PENDING')$
\item for all $\OPH \in \DOM(P)$, if  $\PENDING (\OPH) = \Angle{\OP,
    \TIME, \STATUS}$, then either
  \begin{itemize}
  \item     $\PENDING'    (\OPH) = \PENDING (\OPH)$; or
  \item $\STATUS  = \STATUSPENDING$ and $\PENDING' (\OPH) =
    \Angle{\OP, \TIME, \STATUSTIMEOUT}$; or
  \item  $\STATUS  = \STATUSPENDING$, $\BLOCKCHAIN'.\TIME - \TIME
    \le 60 $, and
    $\PENDING' (\OPH) =       \Angle{\OP, \TIME, \STATUSINCLUDED\
      \BLOCKCHAIN.\TIME}$.
  \end{itemize}

 \item $ \DOM (\MANAGERS) \subseteq \DOM (\MANAGERS')$
 \item for all $\PUK \in \DOM (\MANAGERS)$ if $\MANAGERS (\PUK) =
   \Angle{\BAL, \COU}$ where $\COU = (n, b)$
   \begin{itemize}
   \item if $b= \FALSE$, then $n' = n$
   \item if $b= \TRUE$ and $b'=\TRUE$, then $n' =n$
   \item if $b=\TRUE$ and $b'=\FALSE$, then $n' \in \{ n,  n+1\}$
   \end{itemize}

 \item $ \DOM (\CONTRACTORS) \subseteq \DOM (\CONTRACTORS')$

\begin{definition}%[Blockchain system]
A \emph{blockchain configuration} is a pair
$ \NODE \| \BLOCKCHAIN$ where
\fxnote[inline]{Should the model contain multiple nodes?}
\begin{enumerate}
\item $\NODE = [\EXPRS, \ACCOUNTS, \OPERATIONS]$ is the state of a node, and
\item $\BLOCKCHAIN = [\PENDING, \MANAGERS, \CONTRACTORS,
  \TIME]$ is the state of a blockchain  
  such that
  \begin{itemize}
  \item
    $\forall c \in \ACCOUNTS \implies \exists k\in \MANAGERS, k.\PUK =
    c.\PUK$;
  \item if $\PENDING (\OPH) = \Angle{\OP,\TIME, \STATUSINCLUDED\ \TIME'}$,
    then $\TIME' > \TIME$
  \end{itemize}
\end{enumerate}
\end{definition}
